Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-01-09 analysis/paper.tex
Correct and refactor archaeo_pred

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 81 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. Modelled ranges do not
reliably predict the observed occurrence of specific species at
archaeological sites, however {[}\ldots{]} The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

\begin{itemize}
\tightlist
\item[$\boxtimes$]
  Stop modelling
\item[$\square$]
  First draft:

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    Introduction
  \item[$\square$]
    Background

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\boxtimes$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials
  \item[$\boxtimes$]
    Results
  \item[$\boxtimes$]
    Discussion
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\boxtimes$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{cite}. Even before the end of the Ice Age, these
species supported uniquely dense and complex Late Epipalaeolithic
(15--11.7 ka) societies based on intensive foraging \citep{cite} and
eventually pre-domestication cultivation \citep{cite}. As they were
brought under domestication in the Pre-Pottery Neolithic period
(11.7--8.5 ka), the world's first agriculture was shaped by the
ecosystems from which it emerged and was embedded in.

@@

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the range of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguihed from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworskyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic site of XX \citep{CollinsEtAl2018}. Modelling the spread of
crops has been another significant archaeological application
\citep{CremaEtAl}.

@@

The alternative approach is to train the model using contemporary
occurrence and environmental data and then use palaeoenvironmental data
to `hindcast' its predictions backwards in time. Like
\citet{FranklinEtAl2015}, we view the hindcasting approach as more
promising, because training datasets for both occurrences and
environment are far more readily available, complete and reliable for
the present than the past. There is some scepticism in the ecological
niche modelling literature about the ability of such models to make
accurate predictions in unknown environments (like the past)
\citep{CITES}, but here the hindcasting approach also presents an
opportunity: it reserves archaeological occurrence data as an
independent dataset that can be used to assess the retrodictive
performance of the model. This possibily was suggested by
\citet{FranklinEtAl2015} but to our knowledge our study represents the
first attempt to actually do so.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 96
taxa (Table~\ref{tbl-taxa}) comprising the identifiable species observed
at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites,
according to our previous study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset of 117
assemblages collated from three previously published regional
archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
\citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
distinguish between the source of the remains; archaeobotanical
assemblages are subject to a variety of preservational and recovery
biases, so by no means were all the species on our list consumed or even
deliberately collected by people. However, we assume that there presence
at a site of human settlement at least implies that they were part of
the wider ecosystem that supported habitation there.

Taxonomic names were resolved to the canonical form specified in the
GBIF Backbone Taxonomy \citep{GBIFSecretariat2023}. So for example
occurrences for \emph{Bolboschoenus maritimus} also include those
recorded under the older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-taxa}). Domestic species meeting our inclusion criteria
were substituted with their wild progenitor(s), where different.

@@

Georeferenced occurrence data for West Asia was obtained from the Global
Biodiversity Information Facility (GBIF) using via its application
programming interface and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. GBIF was cleaned to removed
fossil occurrences, recorded absences, and records with missing,
imprecise (\textgreater1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small
training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
excluded 81 taxa with less than 50 usable occurrences, following
recommendations for niche models generally and Random Forest-based
models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
as this would have been computationally prohibitive and we were
uncertain what accounted for such a disproportionately high number of
records.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia used to train models (N=78222}

@@

\subsection{Predictor data}\label{predictor-data}

We modelled the occurrence of species as a function of X spatial
predictor variables (Table~\ref{tbl-predictors}). These included:

@@

\begin{table}

\caption{\label{tbl-climate-periods}}

\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr}
\toprule
Period & Age, ka \\ 
\midrule\addlinespace[2.5pt]
Bølling-Allerød (BA) & 14.7–12.9 \\ 
Younger Dryas (YDS) & 12.9–11.7 \\ 
Early Holocene (EH) & 11.7–8.3 \\ 
Current (CUR) & — \\ 
\bottomrule
\end{tabular*}

Palaeoclimatic periods used for hindcasting, after \citet{BrownEtAl2018}

\end{table}%

For hindcasting, we used reconstructed bioclimatic data for 3 key
periods (Table~\ref{tbl-climate-periods}) generated from downscaled
paleoclimate simulations from the HadCM3 general circulation model
\citep{BrownEtAl2018}. Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is not likely that either macroscale topography
or soil characteristics have altered significantly over the period of
time considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our X predictor variables
(Table~\ref{tbl-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' \citep{WrightZiegler2017} and the
`tidymodels' \citep{tidymodels} framework for data preprocessing and
model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (¾) and test (¼) partitions.

@@

We trained Random Forest models for 81 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Table~\ref{tbl-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 324 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (proportion of correctly classified presence and background
samples) ranged between 72\% and 99\%, with an average of 92\%.
Sensitivity (proportion of correctly classified presence samples) ranged
between 79\% and 100\%, with an average of 94\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.978±0.061 Model sensitivity is loosely correlated with the
number of occurrences available for training
(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
models all having less than 300 recorded occurrences: \emph{Avena
fatua}, \emph{Lepidium perfoliatum}, \emph{Triticum turgidum dicoccum},
and \emph{Zygophyllum fabago}. Test metrics and ROC curves for the
individual models are included in the appendix.

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is significantly worse, with only 5\% of
presences in archaeobotanical assemblages successfully predicted.

\begingroup
\setlength\LTleft{0\linewidth}
\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont

\begin{longtable}{@{\extracolsep{\fill}}lrrrrr}

\caption{\label{tbl-taxa}Summary of species modelled}

@@

 & \multicolumn{2}{c}{Occurrences} & \multicolumn{3}{c}{Model} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-6}
Taxon & Archaeological & Present (GBIF) & Accuracy & ROC-AUC & Sensitivity \\ 

@@

{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.99 \\ 
{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.89 \\ 
{\itshape Aegilops speltoides} & — & 701 & 0.93 & 0.98 & 0.94 \\ 
{\itshape Aegilops tauschii} & — & 725 & 0.90 & 0.98 & 0.94 \\ 
{\itshape Aizoanthemopsis hispanica} & 4 & 515 & 0.98 & 0.99 & 0.96 \\ 
{\itshape Ammi majus} & 4 & 900 & 0.99 & 0.99 & 0.98 \\ 
{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.96 & 0.88 \\ 
{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.85 & 0.94 & 0.91 \\ 
{\itshape Arnebia decumbens} & 21 & 324 & 0.94 & 0.98 & 0.93 \\ 
{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Asphodelus aestivus} & 2 & 76 & 0.96 & 0.97 & 0.96 \\ 
{\itshape Atriplex prostrata} & 4 & 266 & 0.97 & 1.00 & 0.97 \\ 
{\itshape Avena fatua} & 2 & 579 & 0.85 & 0.93 & 0.85 \\ 
{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Bolboschoenus maritimus} & 29 & 1040 & 0.98 & 1.00 & 0.99 \\ 
{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 \\ 
{\itshape Bromus sterilis} & 3 & 759 & 0.96 & 1.00 & 0.98 \\ 
{\itshape Buglossoides arvensis} & 23 & 260 & 0.86 & 0.96 & 0.90 \\ 
{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.94 \\ 
{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 \\ 
{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 \\ 
{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.93 \\ 
{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 \\ 
{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.92 \\ 
{\itshape Cicer reticulatum} & 3 & 52 & 0.96 & 1.00 & 0.92 \\ 
{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.95 & 0.89 \\ 
{\itshape Crithopsis delileana} & 3 & 406 & 0.97 & 0.99 & 0.97 \\ 
{\itshape Erodium ciconium} & 2 & 359 & 0.91 & 0.99 & 0.99 \\ 
{\itshape Ficus carica} & 8 & 3495 & 0.93 & 0.98 & 0.93 \\ 
{\itshape Fumaria densiflora} & 4 & 413 & 0.98 & 0.99 & 0.98 \\ 
{\itshape Gypsophila elegans} & 3 & 76 & 0.93 & 0.96 & 0.85 \\ 
{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.99 & 0.95 \\ 
{\itshape Gypsophila vaccaria} & 8 & 511 & 0.96 & 0.99 & 0.93 \\ 
{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.99 & 1.00 & 1.00 \\ 
{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.98 & 0.96 \\ 
{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.98 & 0.99 & 0.97 \\ 
{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 \\ 
{\itshape Hordeum murinum} & 5 & 1949 & 0.95 & 0.99 & 0.96 \\ 
{\itshape Hordeum spontaneum} & 51 & 4656 & 0.96 & 0.99 & 0.97 \\ 
{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.94 \\ 
{\itshape Lathyrus aphaca} & 4 & 2079 & 0.95 & 0.99 & 0.96 \\ 
{\itshape Lathyrus cicera} & 2 & 591 & 0.90 & 0.98 & 0.94 \\ 
{\itshape Lathyrus oleraceus} & 6 & 1580 & 0.88 & 0.97 & 0.94 \\ 
{\itshape Lathyrus sativus} & 4 & 196 & 0.80 & 0.94 & 0.87 \\ 
{\itshape Lepidium perfoliatum} & 3 & 83 & 0.80 & 0.92 & 0.83 \\ 
{\itshape Linum bienne} & 9 & 376 & 0.96 & 1.00 & 0.98 \\ 
{\itshape Lolium rigidum} & 5 & 2529 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Lolium temulentum} & 3 & 159 & 0.93 & 0.95 & 0.92 \\ 
{\itshape Medicago astroites} & 15 & 143 & 0.90 & 0.97 & 0.88 \\ 
{\itshape Medicago minima} & 2 & 1379 & 0.92 & 0.99 & 0.98 \\ 
{\itshape Medicago radiata} & 20 & 799 & 0.93 & 0.99 & 0.97 \\ 
{\itshape Moltkia coerulea} & 2 & 71 & 0.86 & 0.99 & 1.00 \\ 
{\itshape Neotorularia torulosa} & 2 & 539 & 0.98 & 0.99 & 0.96 \\ 
{\itshape Peganum harmala} & 2 & 917 & 0.91 & 0.98 & 0.95 \\ 
{\itshape Phalaris paradoxa} & 3 & 1191 & 0.99 & 0.99 & 0.98 \\ 
{\itshape Phragmites australis} & 4 & 3412 & 0.97 & 0.99 & 0.97 \\ 
{\itshape Pistacia atlantica} & 6 & 1874 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Plantago lagopus} & 2 & 1730 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Poa bulbosa} & 5 & 1880 & 0.95 & 0.98 & 0.90 \\ 

>>>>

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. Modelled ranges do not
reliably predict the observed occurrence of specific species at
archaeological sites, however {[}\ldots{]} The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{cite}. These species supported uniquely dense and
complex Late Epipalaeolithic (15--11.7 ka) societies based on intensive
foraging \citep{cite} and eventually pre-domestication cultivation
\citep{cite}. As they were brought under domestication in the
Pre-Pottery Neolithic period (11.7--8.5 ka), the world's first
agriculture was shaped by the ecosystems from which it emerged and was
embedded in.

@@

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the range of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---or more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguihed from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworskyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic sites on the Konya Plain \citep{CollinsEtAl2018}. Modelling
the spread of crops has been another significant archaeological
application \citep[e.g.][]{CremaEtAl}, though not as yet applied to West
Asia.

@@

The alternative approach is to train the model using contemporary
occurrence and environmental data and then use palaeoenvironmental data
to `hindcast' its predictions backwards in time. Like
\citet{FranklinEtAl2015}, we view the hindcasting approach as more
promising, because training datasets for both occurrences and
environment are far more readily available, complete and reliable for
the present than the past. There is some scepticism in the ecological
niche modelling literature about the ability of such models to make
accurate predictions in unknown environments \citep[like the
past,][]{FranklinEtAl2015}, but here the hindcasting approach also
presents an opportunity: it reserves archaeological occurrence data as
an independent dataset that can be used to assess the retrodictive
performance of the model. This possibily was suggested by
\citet{FranklinEtAl2015} but to our knowledge our study represents the
first attempt to actually do so.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 97
taxa (Table~\ref{tbl-results-summary}) comprising the identifiable
species observed at at least two Late Epipalaeolithic/Pre-Pottery
Neolithic sites, according to our previous study of the regional
archaeobotanical data \citep{ArranzOtaeguiRoe2023}. This was based on
dataset collated from three previously published regional
archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
\citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

Taxonomic names were resolved to the canonical form specified in the
GBIF Backbone Taxonomy \citep{GBIFSecretariat2023}. So for example
occurrences for \emph{Bolboschoenus maritimus} also include those
recorded under the older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where
different.

@@

Georeferenced occurrence data for West Asia was obtained from the Global
Biodiversity Information Facility (GBIF) using via its application
programming interface and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. GBIF was cleaned to removed
fossil occurrences, recorded absences, and records with missing,
imprecise (\textgreater1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small
training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
excluded 14 taxa with less than 50 usable occurrences, following
recommendations for niche models generally and Random Forest-based
models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
as this would have been computationally prohibitive and we were
uncertain what accounted for such a disproportionately high number of
records.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia used to train models (N=81641}

@@

\subsection{Predictor data}\label{sec-predictors}

We modelled the occurrence of species as a function of 24 geospatial
predictor variables. These included:

@@

For hindcasting, we used reconstructed bioclimatic data for three key
climatologies generated from downscaled paleoclimate simulations from
the HadCM3 general circulation model
\citep{FordhamEtAl2017, BrownEtAl2018}: the Bølling--Allerød
(c.~14.7--12.9 ka), the Younger Dryas (c.~12.9--11.7 ka), and the Early
Holocene (11.7--8.3 ka). Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is unlikely that either macroscale topography or
soil characteristics have altered significantly over the period of time
considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our 24 predictor variables
(Section~\ref{sec-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' \citep{WrightZiegler2017} and the
`tidymodels' \citep{tidymodels} framework for data preprocessing and
model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (¾) and test (¼) partitions.

@@

We trained Random Forest models for 82 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Section~\ref{sec-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 328 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition
Table~\ref{tbl-results-summary}. Model accuracy (proportion of correctly
classified presence and background samples) ranged between 73\% and
99\%, with an average of 93\%. Sensitivity (proportion of correctly
classified presence samples) ranged between 79\% and 100\%, with an
average of 94\%. The area under the models' receiver operating
characteristic curves (ROC-AUC) was on average 0.978±0.061 Model
sensitivity is loosely correlated with the number of occurrences
available for training (Figure~\ref{fig-model-sensitivity-vs-n_occ}),
with the worst-performing models all having less than 300 recorded
occurrences: \emph{Asphodelus aestivus}, \emph{Triticum turgidum
dicoccum}, \emph{Zygophyllum fabago}, and \emph{Stipa dregeana}. Test
metrics and ROC curves for the individual models are included in the
appendix.

\begin{verbatim}
Joining with `by = join_by(taxon)`
\end{verbatim}

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is worse, with 10\% of presences in
archaeobotanical assemblages successfully predicted. Model sensitivity
(proportion of corrected predicted presenses) in relation to the
archaeological data is on average 0.06±0.13
Table~\ref{tbl-results-summary}. A full assessment of the hindcasting
performance of the individual models can be found in the appendix.

\begingroup
\setlength\LTleft{0\linewidth}
\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
\setlength{\LTpost}{0mm}

\begin{longtable}{@{\extracolsep{\fill}}lrrrrrr}

\caption{\label{tbl-results-summary}Summary of species modelled}

@@

 & \multicolumn{2}{c}{Occurrences} & \multicolumn{4}{c}{Model} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-7}
Taxon & Arch. & Cur. & Acc. & ROC-AUC & Sens. & Sens. (arch.) \\ 

@@

{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.86 & 0.00 \\ 
{\itshape Aegilops speltoides} & — & 701 & 0.92 & 0.98 & 0.95 & — \\ 
{\itshape Aegilops tauschii} & — & 725 & 0.89 & 0.98 & 0.94 & — \\ 
{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 515 & 0.98 & 1.00 & 0.96 & 0.00 \\ 
{\itshape Ammi majus} & 4 & 900 & 0.98 & 0.99 & 0.97 & 0.00 \\ 
{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.95 & 0.88 & 0.12 \\ 
{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.84 & 0.95 & 0.91 & 0.00 \\ 
{\itshape Arnebia decumbens} & 21 & 324 & 0.95 & 0.95 & 0.89 & 0.00 \\ 
{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 0.99 & 0.94 & 0.00 \\ 
{\itshape Asphodelus aestivus} & 2 & 76 & 0.97 & 0.97 & 0.83 & 0.00 \\ 
{\itshape Atriplex prostrata} & 4 & 266 & 0.98 & 1.00 & 0.94 & 0.00 \\ 
{\itshape Avena fatua} & 2 & 579 & 0.86 & 0.93 & 0.86 & 0.00 \\ 
{\itshape Avena sterilis} & 5 & 53244\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
{\itshape Bassia prostrata} & 2 & 46\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Bolboschoenus glaucus} & 5 & 27\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{3}} & 31 & 1040 & 0.98 & 0.99 & 0.97 & 0.03 \\ 
{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 & 0.00 \\ 
{\itshape Bromus sterilis} & 3 & 759 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
{\itshape Buglossoides arvensis} & 23 & 260 & 0.87 & 0.97 & 0.95 & 0.09 \\ 
{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.95 & 0.08 \\ 
{\itshape Camphorosma monspeliaca} & 2 & 38\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 & 0.00 \\ 
{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.94 & 0.00 \\ 
{\itshape Celtis tournefortii} & 2 & 14\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 & 0.50 \\ 
{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.94 & 0.00 \\ 
{\itshape Cicer reticulatum}\textsuperscript{\textit{4}} & 3 & 52 & 0.97 & 1.00 & 1.00 & 0.00 \\ 
{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.96 & 0.90 & 0.00 \\ 
{\itshape Crithopsis delileana} & 3 & 406 & 0.98 & 1.00 & 0.97 & 0.00 \\ 
{\itshape Erodium ciconium} & 2 & 359 & 0.92 & 0.99 & 0.94 & 0.00 \\ 
{\itshape Euclidium syriacum} & 3 & 32\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Ficus carica} & 8 & 3495 & 0.94 & 0.99 & 0.95 & 0.25 \\ 
{\itshape Fumaria densiflora} & 4 & 413 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
{\itshape Gypsophila elegans} & 3 & 76 & 0.92 & 0.99 & 1.00 & 0.00 \\ 
{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.97 & 0.90 & 0.00 \\ 
{\itshape Gypsophila vaccaria}\textsuperscript{\textit{5}} & 8 & 511 & 0.96 & 0.99 & 0.93 & 0.00 \\ 
{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
{\itshape Henrardia pubescens} & 3 & 1\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 & 0.00 \\ 
{\itshape Hordeum murinum} & 5 & 1949 & 0.94 & 0.98 & 0.93 & 0.00 \\ 
{\itshape Hordeum spontaneum}\textsuperscript{\textit{6}} & 110 & 4656 & 0.96 & 0.99 & 0.97 & 0.20 \\ 
{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.95 & 0.00 \\ 
{\itshape Krascheninnikovia ceratoides} & 2 & 42\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Lathyrus aphaca} & 4 & 2079 & 0.94 & 0.99 & 0.97 & 0.00 \\ 
{\itshape Lathyrus cicera} & 2 & 591 & 0.89 & 0.98 & 0.94 & 0.50 \\ 
